dyngen: benchmarking with in silico single cells

### 2019-12-01 manuscript.Rmd
update manuscript

**Purpose:** A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.  
**Results:** We developed `dyngen`, a multi-modality simulator of single cells. In `dyngen`, the biomolecular state of an _in silico_ cell changes over time according to a set of reactions defined by the cell's gene regulatory network. By simulating single cells in terms of its biomolecular state and reactions, the simulator is easily extendible for adding new modalities or experimental procedures. We demonstrate `dyngen`'s flexibility by simulating snapshot, time-series and perturbation experiments.  
**Conclusion:** `dyngen` lays the foundations for benchmarking a wide variety of computational single-cell tools and can be used to help kick-start the development of future types of analyses.

@@

We introduce `dyngen`, a multi-modality simulator of single cells and their dynamics (Figure\ \ref{fig:showcase}A). A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] where a cell consists of a set of molecules. Throughout a simulation, reactions (e.g. transcription, splicing) modify the abundance of these molecules. The likelihood of a reaction occurring is again dependent on molecule abundance.

By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure\ \ref{fig:showcase}B). We demonstrate `dyngen`'s flexibility by simulating snapshot, time-series and perturbation experiments, allowing to benchmark a wide variety of computational tools such as trajectory alignment and differential network inference methods.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=\linewidth]{fig/showcase_4}
    \caption{
            \textbf{Showcase of \texttt{dyngen} functionality.}
            \textbf{A:}
            \textbf{B:} Evaluating different types of computational tools requires simulating different types of experiments and extracting different layers of information from the simulation.
    }
    \label{fig:showcase}
\end{figure}

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=\linewidth]{fig/simplecyclic_edited}
    \caption{
            \textbf{Key features of \texttt{dyngen} are illustrated using a cyclic toy example.}
            While this example comprises only of a single cell and 5 genes, \texttt{dyngen} is able to simulate thousands of cells with GRNs containing thousands of genes.
    }
    \label{fig:simplecyclic}
\end{figure}

`dyngen` returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure\ \ref{fig:simplecyclic}C--F). These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches. For example, a network inference method could use mRNA abundance and cellular states as inputs and its output could be benchmarked against the gold standard GRN.

Depending on how the GRN is designed, different cellular developmental processes can be simulated.
`dyngen` includes generators of GRNs which result in many different developmental topologies (Figure\ \ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.6\textwidth]{fig/example_runs_2}
    \caption{
            \textbf{Multiple executions of \texttt{dyngen} with different predefined backbones.} From each simulation of about 200 genes, 1000 cells were sampled.
            \textbf{A:}~Linear. \textbf{B:}~Bifurcating. \textbf{C:}~Converging.
            \textbf{D:}~Cyclic. \textbf{E:}~Bifurcating loop. \textbf{F:}~Bifurcating converging.
            \textbf{G:}~Consecutive branching. \textbf{H:}~Binary tree. \textbf{I:}~Disconnected.
    }
    \label{fig:example_runs}
\end{figure}

Custom-defined GRNs offer more fine-grained control over the simulation.
Aside from simulating topologies currently not supported by `dyngen`, this has several other important use-cases. Simulations of the same GRN with small perturbations allow to emulate batch effects or perturbation experiments (Figure\ \ref{fig:usecases}).
Simulating perturbed GRNs allow evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other -- or differential network inference methods -- which infer differential regulatory interactions between two or more groups of profiles.

@@

`dyngen` can be used to simulate different experimental conditions. By default, `dyngen` supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
However, it is possible to imagine and implement other sampling strategies, such as sampling a cell at a certain time point and once more at a later time point. This would allow evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.

<!-- YSA: RNA velocity: Onduidelijk hier, mensen kennen dit niet. -->

@@

# Discussion {#sec:dyngen-discussion}
As is, `dyngen`'s single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications. This may include methods that integrate clustering, network inference and trajectory inference. In this respect, `dyngen` may promote the development of new tools in the single-cell field similarly as other simulators have done in the past[@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].

`dyngen` ultimately allows anticipating technological developments in single-cell multi-omics. In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
In addition, it could also be used to compare which experimental protocol is the most cost-effective in producing the qualitative and robust results in downstream analysis.

Currently, `dyngen` focuses on simulating cells as standalone entities that are well mixed.
Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration[@smith_spatialstochasticintracellular_2019].

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=\textwidth]{fig/explain_methods}
    \caption{\textbf{The workflow of \texttt{dyngen} is comprised of six main steps.} 
    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. 
    \textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module. Additional target genes are generated. 
    \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. 
    \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states. The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge. 
    \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.  The counts of each of the molecules at each time step are extracted. Each time step is mapped to a point in the backbone. 
    \textbf{F:} Multiple cells are sampled from each simulation. Molecules are sampled from each cell.
    }
    \label{fig:explain_methods}
\end{figure}

@@

In `dyngen`, the user defines the behaviour of the simulation by defining how sets of genes, called modules, are regulating each other.
A module may have basal expression, which means that pre-mRNA of the genes in this module will be transcribed without the presence of transcription factor molecules. A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase. At the end of the `dyngen` process, cells will not be sampled from the burn phase simulations. Interactions between modules have a strength (which is a positive integer) and an effect (+1 for upregulating, -1 for downregulating).

@@

Note that these examples represent the bare minimum in terms of number of modules used. Using longer chains of modules is typically desired. In addition, the fate decisions made in this example of a bifurcation is reversible, meaning cells can be reprogrammed to go down a different differentiation path. If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring.

@@

### Backbone lego
The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure\ \ref{fig:backbone_lego}). A BBL consists of one or more modules which regulate each other such that the output modules present a specific behaviour, depending on the input module (Figure\ \ref{fig:backbone_lego}A). Parameters allow determining the number of modules involved in the process and the number of outputs. Multiple BBLs can be chained together in order to intuitively create milestone networks and corresponding state networks (Figure\ \ref{fig:backbone_lego}B). Note that not all dynamic processes can be represented by a combination of BBLs, but they can serve as common building blocks to aid the construction of the backbone.

@@

<!-- paper on transcriptional bursting -->

<!-- doi: 10.1039/C7MB00154A -->

<!-- interesting to try to replicate in dyngen! :) -->

@@

\newcommand{\wpr}[1]{\text{wpr}_{#1}}
\newcommand{\wsr}[1]{\text{wsr}_{#1}}
\newcommand{\xdr}[1]{\text{xdr}_{#1}}
\newcommand{\ypr}[1]{\text{ypr}_{#1}}
\newcommand{\ydr}[1]{\text{ydr}_{#1}}

\newcommand{\str}[1]{\text{str}_{#1}}
\newcommand{\hill}[1]{\text{hill}_{#1}}
\newcommand{\coop}[1]{\text{coop}_{#1}}
\newcommand{\dis}[1]{\text{dis}_{#1}}
\newcommand{\buf}[1]{\chi_{#1}}
\newcommand{\ba}[1]{\text{ba}_{#1}}

@@

Throughout a simulation, the abundance levels of molecules are affected by five reactions, namely transcription, splicing, mRNA degradation, translation, and protein degradation. Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered. The effects and propensity functions of these reactions are defined in Table\ \ref{tab:reaction_def}.

<!-- TODO: add step-by-step  derivation of transcription formula -->

<!-- YSA: Graag wat meer uitleg in de tekst over al deze formules, definitie van notaties en zo. -->

\begin{table}[ht]
    \caption{
      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
    } \label{tab:reaction_def}
    \centering
    \begin{tabular}{|lrl|}
            \hline
            Parameter & Symbol & Definition \\ \hline \hline
            Transcription rate & $\wpr{G}$ & $\in N(100, 20),\ \geq 10$ \\
            Splicing rate & $\wsr G$ & $\in N(10, 2),\ \geq 2$ \\
            mRNA degradation rate & $\xdr{G}$ & $\in N(5, 1),\ \geq 2$ \\
            Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 2$ \\
            Protein degradation rate & $\ydr G$ & $\in N(3, 0.5),\ \geq 1$ \\
            Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
            Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
        Cooperativity factor & $\coop G$ & $\in [0, 1]$ * \\ \hline\hline
            Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\
            Regulation activity & $\buf{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\dis H}\right) ^ {\hill{G,H}}$ \\
            Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
            \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
    \end{tabular}
\end{table}

<!-- 
% regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]). y_H / maxy_H is more likely to be a probability that H is bound to G, though this does not take into account that H can be bound to other genes as well.
% in any case, if it is, then it upregulates G by strG,H. It is assumed binding events
% are independent and result in multiplicative effect.

% according to https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)#Proportion_of_ligand-bound_receptors:
% P_H = (y_H ^ N) / (degradation_H / production_H + y_H ^ N)
% ...
% ...
% might need to go back to things like:
% https://kar.kent.ac.uk/24077/1/myzabetpaper.pdf
-->

\begin{table}[ht]
    \caption{
      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
    } \label{tab:reaction_params}
    \centering
    \begin{tabular}{|lrl|}
            \hline
            Parameter & Symbol & Definition \\ \hline \hline
            Transcription rate & $\wpr{G}$ & $\in N(100, 20),\ \geq 10$ \\
            Splicing rate & $\wsr G$ & $\in N(10, 2),\ \geq 2$ \\
            mRNA degradation rate & $\xdr{G}$ & $\in N(5, 1),\ \geq 2$ \\
            Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 2$ \\
            Protein degradation rate & $\ydr G$ & $\in N(3, 0.5),\ \geq 1$ \\
            Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
            Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
        Cooperativity factor & $\coop G$ & $\in [0, 1]$ * \\ \hline\hline
            Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\
            Regulation activity & $\buf{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\dis H}\right) ^ {\hill{G,H}}$ \\
            Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
            \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
    \end{tabular}
\end{table}

@@

<!--  mention that this is basically an ode? -->
## Simulate single cells {#sec:dyngen-simcell}

<!-- 
%TODO: STOCKS[@kierzek_stocksstochastickinetic_2002] uses GSSA to simulate something.
% maybe we could have a look at that!
% or https://nyaspubs.onlinelibrary.wiley.com/doi/full/10.1111/j.1749-6632.2008.03756.x

% or even https://academic.oup.com/bioinformatics/article/24/10/1318/178687 ('AND' and 'OR')
% and https://academic.oup.com/bioinformatics/article/25/9/1205/203870 (regulation motifs, kinetics)
% https://nyaspubs.onlinelibrary.wiley.com/doi/full/10.1111/j.1749-6632.2008.03756.x (modules of genes)

%TODO: improve this section a lot. Basically, explain SSA and GillespieSSA2.
-->

`dyngen` uses Gillespie's stochastic simulation algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.

@@

`GillespieSSA2` is an optimised library for performing SSA simulations. The propensity functions are compiled to C++ and SSA approximations can be used which allow to trigger many reactions simultaneously at each iteration. The framework also allows to store the abundance levels of molecules only after a specific interval has passed since the previous census. By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information. In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained, as well as specific sub-calculations of the propensity values, such as the regulator activity level $reg_{G,H}$.

@@

## Simulate experiment {#sec:dyngen-experiment}
From the SSA simulation we obtain the abundance levels of all the molecules at the different time points. We need to replicate technical effects introduced by experimental protocols in order to obtain data that is similar to real data. For this, the cells are sampled from the simulations and molecules are sampled for each of the cells. Real datasets are used in order to achieve similar data characteristics.

### Sample cells
Cells can be sampled from an unsynchronised population of single cells (snapshot) or at multiple time points in a synchronised population (time series).

**Snapshot:** Cells are sampled uniformly from the different time points in the simulation.

**Time series:** The timeline of the simulations is divided into multiple chunks. From several of these chunks, cells are sampled. For each cell it is known at which time point it was sampled.

@@

# Author contributions {#sec:dyngen-contributions}

* R.C. and W.S. designed the study.
* R.C. and W.S. performed the experiments and analysed the data.
* R.C. and W.S. implemented the `dyngen` software package.
* R.C. wrote the original manuscript.
* R.C., W.S., and Y.S. reviewed and edited the manuscript.
* Y.S. supervised the project.

>>>>

**Purpose:** A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.  
**Results:** We developed dyngen, a multi-modality simulator of single cells. In dyngen, the biomolecular state of an _in silico_ cell changes over time according to a set of reactions defined by the cell's gene regulatory network. By simulating single cells in terms of its biomolecular state and reactions, the simulator is easily extendible for adding new modalities or experimental procedures. We demonstrate dyngen's flexibility by simulating snapshot, time-series and perturbation experiments.  
**Conclusion:** dyngen lays the foundations for benchmarking a wide variety of computational single-cell tools and can be used to help kick-start the development of future types of analyses.

**Author contributions:**

* R.C. and W.S. designed the study.
* R.C., W.S., and L.D. performed the experiments and analysed the data.
* R.C. and W.S. implemented the dyngen software package.
* R.C. wrote the original manuscript.
* R.C., W.S., L.D., and Y.S. reviewed and edited the manuscript.
* Y.S. supervised the project.

@@

We introduce dyngen, a multi-modality simulator of single cells and their dynamics (Figure\ \ref{fig:showcase}A). A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] where a cell consists of a set of molecules. Throughout a simulation, reactions (e.g. transcription, splicing) modify the abundance of these molecules. The likelihood of a reaction occurring is again dependent on molecule abundance.

By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure\ \ref{fig:showcase}B). We demonstrate dyngen's flexibility by simulating snapshot, time-series and perturbation experiments, allowing to benchmark a wide variety of computational tools such as trajectory alignment and differential network inference methods.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=\linewidth]{fig/showcase_4}
    \caption{
        \textbf{Showcase of dyngen functionality.}
        \textbf{A:}
        \textbf{B:} Evaluating different types of computational tools requires simulating different types of experiments and extracting different layers of information from the simulation.
    }
    \label{fig:showcase}
\end{figure}

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=\linewidth]{fig/simplecyclic_edited}
    \caption{
        \textbf{Key features of dyngen are illustrated using a cyclic toy example.}
        While this example comprises only of a single cell and 5 genes, dyngen is able to simulate thousands of cells with GRNs containing thousands of genes.
    }
    \label{fig:simplecyclic}
\end{figure}

dyngen returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure\ \ref{fig:simplecyclic}C--F). These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches. For example, a network inference method could use mRNA abundance and cellular states as inputs and its output could be benchmarked against the gold standard GRN.

Depending on how the GRN is designed, different cellular developmental processes can be simulated.
dyngen includes generators of GRNs which result in many different developmental topologies (Figure\ \ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.6\textwidth]{fig/example_runs_2}
    \caption{
        \textbf{Multiple executions of dyngen with different predefined backbones.} From each simulation of about 200 genes, 1000 cells were sampled.
        \textbf{A:}~Linear. \textbf{B:}~Bifurcating. \textbf{C:}~Converging.
        \textbf{D:}~Cyclic. \textbf{E:}~Bifurcating loop. \textbf{F:}~Bifurcating converging.
        \textbf{G:}~Consecutive branching. \textbf{H:}~Binary tree. \textbf{I:}~Disconnected.
    }
    \label{fig:example_runs}
\end{figure}

Custom-defined GRNs offer more fine-grained control over the simulation.
Aside from simulating topologies currently not supported by dyngen, this has several other important use-cases. Simulations of the same GRN with small perturbations allow to emulate batch effects or perturbation experiments (Figure\ \ref{fig:usecases}).
Simulating perturbed GRNs allow evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other -- or differential network inference methods -- which infer differential regulatory interactions between two or more groups of profiles.

@@

dyngen can be used to simulate different experimental conditions. By default, dyngen supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
However, it is possible to imagine and implement other sampling strategies, such as sampling a cell at a certain time point and once more at a later time point. This would allow evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.

@@

# Discussion {#sec:dyngen-discussion}
As is, dyngen's single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications. This may include methods that integrate clustering, network inference and trajectory inference. In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past[@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].

dyngen ultimately allows anticipating technological developments in single-cell multi-omics. In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
In addition, it could also be used to compare which experimental protocol is the most cost-effective in producing the qualitative and robust results in downstream analysis.

Currently, dyngen focuses on simulating cells as standalone entities that are well mixed.
Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration[@smith_spatialstochasticintracellular_2019].

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=\textwidth]{fig/explain_methods}
    \caption{\textbf{The workflow of dyngen is comprised of six main steps.} 
    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. 
    \textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module. Additional target genes are generated. 
    \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. 
    \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states. The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge. 
    \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.  The counts of each of the molecules at each time step are extracted. Each time step is mapped to a point in the backbone. 
    \textbf{F:} The molecule levels of multiple simulations are shown over time (left). From each simulation, multiple cells are sampled (from left to middle). Technical noise from profiling is simulated by sampling molecules from the set of molecules inside each cell (from middle to right).
    }
    \label{fig:explain_methods}
\end{figure}

@@

To start a dyngen simulation, the user needs to define a module network and a backbone. The module network defines how sets of co-regulated genes, called modules, regulate each other. The module network is what mainly determines which dynamic processes occur within the simulated cells. The backbone is a separate set of simulations in which the ground-truth topology of the dynamic processes are defined, as it is difficult to determine the topology of the dynamic processes from the module network itself.

A module network consists of modules connected together by regulatory interactions. 
A module may have basal expression, which means genes in this module will be transcribed without the presence of transcription factor molecules. A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen-simcell}). At the end of the dyngen process, cells will not be sampled from the burn phase simulations. Interactions between modules have a strength (which is a positive integer) and an effect (+1 for upregulating, -1 for downregulating).

@@

Note that these examples represent the bare minimum in terms of number of modules used. Using longer chains of modules is typically desired. In addition, the fate decisions made in this example of a bifurcation is reversible, meaning cells can be reprogrammed to go down a different differentiation path. If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring (Section \ref{sec:dyngen-bbl}).

@@

### Backbone lego {#sec:dyngen-bbl}
The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure\ \ref{fig:backbone_lego}). A BBL consists of one or more modules which regulate each other such that the output modules present a specific behaviour, depending on the input module (Figure\ \ref{fig:backbone_lego}A). Parameters allow determining the number of modules involved in the process and the number of outputs. Multiple BBLs can be chained together in order to intuitively create module networks and corresponding state networks (Figure\ \ref{fig:backbone_lego}B). Note that not all dynamic processes can be represented by a combination of BBLs, but they can serve as common building blocks to aid the construction of the backbone.

@@

@@

\newcommand{\wpr}[1]{\text{wpr}_{#1}}
\newcommand{\whl}[1]{\text{whl}_{#1}}
\newcommand{\wdr}[1]{\text{wdr}_{#1}}
\newcommand{\wsr}[1]{\text{wsr}_{#1}}
\newcommand{\xhl}[1]{\text{xhl}_{#1}}
\newcommand{\xdr}[1]{\text{xdr}_{#1}}
\newcommand{\ypr}[1]{\text{ypr}_{#1}}
\newcommand{\yhl}[1]{\text{yhl}_{#1}}
\newcommand{\ydr}[1]{\text{ydr}_{#1}}

\newcommand{\str}[1]{\text{str}_{#1}}
\newcommand{\hill}[1]{\text{hill}_{#1}}
\newcommand{\coop}[1]{\text{co}_{#1}}
\newcommand{\dis}[1]{\text{dis}_{#1}}
\newcommand{\buf}[1]{\chi_{#1}}
\newcommand{\ba}[1]{\text{ba}_{#1}}

@@

Throughout a simulation, the abundance levels of molecules are affected by five reactions, namely transcription, splicing, mRNA degradation, translation, and protein degradation. Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.

The effects of each reaction mimic the respective biological processes (Table\ \ref{tab:reaction_def}, middle). Transcription of gene $G$ results in the creation of a single pre-mRNA molecule $\w G$. Splicing turns one pre-mRNA $\w G$ into a mature mRNA $\w G$. Translation uses a mature mRNA $\x G$ to produce a protein $\y G$. Pre-mRNA, mRNA and protein degradation results in the removal of a $\w G$, $\x G$, and $\y G$ molecule, respectively.

The propensity of all reactions except transcription are straight-forward (Table\ \ref{tab:reaction_def}, right). The degradation and splicing propensity of pre-mRNAs of a gene $G$ are linear functions of the abundance $\w G$, namely $\wdr G \times \w G$ and $\wsr G \times \w G$ respectively. Similarly, the degradation and translation of mature mRNAs are linear functions of $\x G$, namely $\xdr G \times \x G$ and $\ypr G \times \x G$. Degradation of proteins is also a linear function of $\y G$, namely $\ydr G \times \y G$. Constant parameters $\wdr G$, $\wsr G$, $\xdr G$, $\ypr G$, and $\ydr G$ are from respective normal distributions (Table\ \ref{tab:reaction_params}). These formulae are based on those of GNW[@schaffter_genenetweaversilicobenchmark_2011], except that the production and decay of molecules have been separated into two separate formulae.

\begin{table}[ht]
    \caption{
      \textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mature mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$. Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
    } \label{tab:reaction_def}
    \centering
    \begin{tabular}{|lcc|}
        \hline
        Reaction & Effect & Propensity \\ \hline \hline
        Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \coop{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\coop G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
        Pre-mRNA degradation & $\w G \rightarrow \emptyset$ & $\wdr G \times \w G$ \\
        Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
        Mature mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
        Translation & $\x G \rightarrow \x G + \y G$ & $\ypr G \times \x G$ \\
        Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
    \end{tabular}
\end{table}

\begin{table}[ht]
    \caption{
      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
    } \label{tab:reaction_params}
    \centering
    \begin{tabular}{|lrl|}
        \hline
        Parameter & Symbol & Definition \\ \hline \hline
        Transcription rate & $\wpr{G}$ & $\in N(50, 10),\ \geq 10$ \\
        Splicing rate & $\wsr G$ & $\in N(5, 1),\ \geq 1$ \\
        Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 1$ \\
        Pre-mRNA half-life & $\whl{G}$ & $\in N(0.15, 0.03),\ \geq 0.05$ \\
        Mature mRNA half-life & $\xhl{G}$ & $\in N(0.15, 0.03),\ \geq 0.05$ \\
        Protein half-life rate & $\yhl G$ & $\in N(0.25, 0.05),\ \geq 0.1$ \\
        Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
        Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
        Cooperativity factor & $\coop G$ & $\in [0, 1]$ * \\ \hline\hline
        Pre-mRNA degradation rate & $\wdr G$ & $= \ln(2)\ /\ \whl G$ \\
        Mature mRNA degradation rate & $\xdr G$ & $= \ln(2)\ /\ \xhl G$ \\
        Protein degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\wdr H \times \xdr H \times \ydr H}$ \\
        Regulatory strength & $\buf{G,H}$ & $= \left(\str{G,H} \times \y H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
        Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
        \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
    \end{tabular}
\end{table}

@@

## Simulate single cells {#sec:dyngen-simcell}
dyngen uses Gillespie's stochastic simulation algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.

@@

GillespieSSA2 is an optimised library for performing SSA simulations. The propensity functions are compiled to C++ and SSA approximations can be used which allow to trigger many reactions simultaneously at each iteration. The framework also allows to store the abundance levels of molecules only after a specific interval has passed since the previous census. By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information. In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained, as well as specific sub-calculations of the propensity values, such as the regulator activity level $reg_{G,H}$.

@@

## Simulate experiment {#sec:dyngen-experiment}
From the SSA simulation we obtain the abundance levels of all the molecules at every state. We need to replicate technical effects introduced by experimental protocols in order to obtain data that is similar to real data. For this, the cells are sampled from the simulations and molecules are sampled for each of the cells. Real datasets are used in order to achieve similar data characteristics.

### Sample cells
In this step, $N$ cells are sampled the simulations. Two approaches are implemented: sampling from an unsynchronised population of single cells (snapshot) or sampling at multiple time points in a synchronised population (time series).

**Snapshot** The backbone consists of several states linked together by transition edges with length $L_i$, to which the different states in the different simulations have been mapped (Figure\ \ref{fig:sample_cells}A). From each transition, $N_i = N / \frac{L_i}{\sum L_i}$ cells are sampled uniformly, rounded such that $\sum N_i = N$.

**Time series** Assuming that the final time of the simulations is $T$, the interval $[0, T]$ is divided into $k$ equal intervals of width $w$ separated by $k-1$ gaps of width $g$. $N_i = N / k$ cells are sampled uniformly from each interval (Figure\ \ref{fig:sample_cells}B), rounded such that $\sum N_i = N$. By default, $k = 8$ and $g = 0.75$. For usual dyngen simulations, $10 \leq T \leq 20$. For larger values of $T$, $k$ and $g$ should be increased accordingly.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=\linewidth]{fig/sample_cells.pdf}
    \caption{
        \textbf{Two approaches can be used to sample cells from simulations: snapshot and time-series.}
    }
    \label{fig:sample_cells}
\end{figure}

@@

## Extract state-specific regulatory effect {#sec:dyngen-extractgrn}
When desired, the regulatory effect of a regulator $R$ on a target $T$ can be determined and returned (See Figure\ \ref{fig:explain_methods}). This information is useful, amongst others, for benchmarking casewise network inference methods. To calculate the regulatory effect of $R$ on $T$ at a particular state during a simulation,

## Comparison of casewise network inference methods {#sec:dyngen:nicompare}
